---
title: "Class 09: Structural Bioinformatics"
author: "Ryan Chung A15848050"
format: html
---

### Section 1 - Intro to RCSB Protein Data Bank (PDB)

```{r}

pdb <- read.csv('Data_Export_Summary.csv')

```

> Q1: What percentage of structures in the PDB are solved by X-Ray and Electron Microscopy.

92.99% of structures in the PDB are solved by X-Ray and Electron Microscopy with each making up approximate 86 and 7% respectively.

```{r}

#xray <- sum(154766, 9083, 8110, 2664, 163, 11)
#em <- sum(10155, 1802, 3176, 94, 9 ,0 )
#total <- sum(177403, 10925, 11575, 4223, 204, 22)

#(xray + em )/ total

 
xtot <- sum(as.numeric(gsub(',', '' , pdb$X.ray)))

etot <- sum(as.numeric(gsub(',','',pdb$EM))) 

ttot <- sum(as.numeric(gsub(',','', pdb$Total)))

(xtot + etot)/ttot

```

Make this a function

```{r}

tot <- function(z) {
  #Take inuput z and remove commas
    #then make it numeric and then take the sum
  #Save the sum as y
  y <- sum(as.numeric(gsub(',', '', z)))
  #return output
  return(y)
}
  
Q1 <- 100* (tot(pdb$X.ray) + tot(pdb$EM)) / tot(pdb$Total)


xprop <- round(tot(pdb$X.ray) / tot(pdb$Total), 2)

eprop <- round(tot(pdb$EM) / tot(pdb$Total), 2)

```

> Q2: What proportion of structures in the PDB are protein?

86.81% of structures in the PDB are protein (only).

```{r}

rmcomma <- function(input){
 input <- as.numeric(gsub(',','', input))
}
  
pdb$Total <- as.numeric(gsub(',', '',pdb$Total))

prot <- sum(pdb$Total[1])

prot/sum(pdb$Total)


```

> Q3: Type HIV in the PDB website search box on the home page and determine how many HIV-1 protease structures are in the current PDB?

There are 239 HIV-1 protease structures in the current PDB.

### Part 2 - Visualizing the HIV-1 protease structure

![HIV-PR Structure from MERK with a bound drug](1HSG(1).png)

> Q4: Water molecules normally have 3 atoms. Why do we see just one atom per water molecule in this structure?

The hydrogen molecules are extremely small and are not rendered due to the low resolution representation,

> Q5: There is a critical "conserved" water molecule in the binding site. Can you identify this water molecule? What residue number does this water molecule have

The molecule is HOH 308.

### Part 3 - Introduction to Bio3D in R

##### Working with Structures in R

We can use the `bio3d` package to read and perform bioinformativcs calculations on PDB structures

```{r}

library(bio3d)

hsg <- read.pdb("1hsg")
hsg
```

```{r}
attributes(hsg)
```


```{r}
head(hsg$atom)
```

#### Part 4 - Comparative structure analysis of Adenylate Kinase

```{r}
adk <- read.pdb('6s36')

```

Perform a prediction of flexibility with a technique called NMA (normal mode analysis)

```{r}
#perform a flexibility prediction
m <- nma(adk)

plot(m)
```

Write out a 'movie' (a.k.a trajectory) of the motion for viewing in MOlstar
```{r}

mktrj(m, file="adk_m7.pdb")

```



>Q7: How many amino acid residues are there in this pdb object? 

198

>Q8: Name one of the two non-protein residues? 

HOH

>Q9: How many protein chains are in this structure? 

2 protein chains













